Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-05-18 MS/ms.Rmd
More writing to hash out ideas and structure a little more...fun, fun!

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance) in a meta-analysis, and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect sizes comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with common treatment differeces across studies; allowing researchers to compare both mean differences (e.g., Q10 to compare mean differences in physiological rates over 10 degrees) and associated differences in variance (e.g, SQ10 for comparing variability in in physiological rates over 10 degrees) for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is a well-developed research field with a long history in the fields of statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis, and is becoming increasingly more prominant the field of comparative physiology. Meta-analysis involves strict literature-based search protocols and data extraction (i.e., following the PRISMA protocol, e.g., [@ODea2021]) to achieve two major goals, 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error []. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on esatimytes, even if the overall mean itself is unbiased.

@@

A very common occurance in comparative physiology is to manipulate the tempearture organisms experience and measure some physiologcal rate (e.g., metabolic rate). Physiological rates are often compared using $Q_{10}$ values (e.g., [@Rodger2021]), which describe the multiplicative change in physiological rates across a 10&deg;C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to weight effect size estimates by sampling variance as is typical in meta-analysis. Recognition of the similarity between *lnRR* and $Q_{10}$ means that we can derive $Q_{10}$ based effect sizes using the well known mathematical properties of *lnRR*, and improve upon our ability to account for sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017]. As such, here we derive a series of $Q_{10}$ based effect sizes along with their associated sampling variances that allow one to compare both changes in mean and variance of physiological rates.

@@

## **Meta-Analytic Models that Control for Treatment Differences Within and Across Studies**

#### *Multi-level Meta-Analysis when the Overall Effect is of Interest*
#### *Heterogenity Analysis: How Much Effect Size Variation Actually Exists?*
Quantifying effect size heterogenity (i.e., effect size variance) provides a pathway to explore hypotheses thought to explain why effect might vary across studies.

#### *Multi-level Metaregression to Understand Variation and Imrpove Interpretation*

>>>>

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance) in a meta-analysis, and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect sizes comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with common treatment differeces across studies; allowing researchers to compare both mean differences (e.g., $Q_{10}$ to compare mean differences in physiological rates over 10 degrees) and associated differences in variance (e.g, $S_{Q10}$ for comparing variability in in physiological rates over 10 degrees) for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is a well-developed research field with a long history in the fields of statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis, and is becoming increasingly more prominant the field of comparative physiology. Meta-analysis involves strict literature-based search protocols and data extraction [i.e., following the PRISMA protocol, see @ODea2021] to achieve two major goals, 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error []. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on esatimytes, even if the overall mean itself is unbiased.

@@

A very common occurance in comparative physiology is to manipulate the tempearture organisms experience and measure some physiologcal rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparablity' issues. As such, it is often of interest to standardsise effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10&deg;C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interstingly, $Q_{10}$ is quite similar to *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analtic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

## **Meta-Analytic Models to Control for Heterogeous Effects Within and Across Studies**

It will not possible to derive an effect size that completely controls for known heterogenity at the effect size level all the time. This may because because treatment effects may not be simple linear or exponential functions of the mean response making the relationship with effect size complex [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with expoential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regerssion approaches, controlling for these known treatment differences as a moderator. There are simple statisical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytical approaches or steps commonly used.

#### *Multi-level Meta-Analysis when the Overall Effect is of Interest*
One of the main goals of meta-analysis is to obtain an overall meta-analytic mean estimate (and some form of uncertainty around that mean, like a 95% confidence interval, may be even with a prediction interval - @Nakagawa2021c) which describes the magnitude of effect across all the studies / effects included in the meta-analysis. This is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierachical structure. For example, multiple effect sizes may be taken from a single study on different traits measured on the same individuals, or come from many different species that share an evolutionary history [@Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multi-level meta-analytic model to control for the various sources of non-independence [@NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance. For the purpose of our discussion, we will assume the meta-analyst plans to use more advanced multilevel models. Having said that, it maybe sufficient for many meta-analyses to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009].

#### *Heterogenity Analysis: How Much Effect Size Variation Actually Exists?*
Quantifying effect size heterogenity (i.e., effect size variance) provides a pathway to explore hypotheses thought to explain why effect might vary across studies.

#### *Multi-level Metaregression to Understand Variation and Improve Interpretation*
